Drug sensitivity prediction with the normal inverse Gaussian prior informed by external data

### 2019-11-25 docs/manuscript1.Rnw
added bSEM to analysis_gdsc.R

	We estimate four models: (i) an NIG model that does include external 
	covariates in the estimation called NIG1 in the following, (ii) the NIG model
	estimated as in Section \ref{sec:estimation} called NIG2, and regular (iii)
	lasso and (iv) ridge models.
	Exclusion of the external covariates as in the NIG1 model
	amounts to direct estimation of one common expected 
	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
	$\phi_{jd}$ and/or $\chi_d$ as in the NIG2 model. All 
	models are fit on 241 randomly selected cell lines, while performance measures 
	are estimated from the remaining 241 cell lines.

@@

	Table \ref{tab:simulations_gdsc_tab1} displays several median performance
	measures for simulation version 1. The table includes as a reference 
	the null model, which
	is an empty model that only fits the mean of the $\y_{id}$.
	The median evidence lower bounds (ELBO) approximates the marginal
	likelihoods of models NIG1 and NIG2 on both training and test data, while 
	estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
	performance. Focussing on MSE, we see that the NIG2 model, that includes
	the external covariates, performs best, both in terms of estimation and 
	prediction. Model fit is best for the NIG1 model that does not include the
	external covariates, according to the ELBO on the training data.
	The lower MSEs confirm that the model that
	includes the external data (NIG2) learns the underlying structure 
	in the data better than model NIG1 that does not include the external 
	data.
<<simulations_gdsc_tab1>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 1, with the best performing model in bold.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

@@

	We expect that drugs that have been clinically approved are easiest to 
	predict and hence yield the largest prior $\beta_{jd}$ variances, followed by
	the drugs in clinical development, and the experimental drugs. Likewise we 
	expect the targeted drugs to yield the largest prior $\beta_{jd}$ variances, 
	followed by the cytotoxic drugs, and the unkown target drugs. For the
	feature covariate, we expect that genes that are in the pathway of the drug
	are more predictive than genes that are not, i.e., 
	they have larger prior $\beta_{jd}$ variances than drugs that are not in 
	the pathway.

	\subsection{Results}
	As a preliminary step, we select
	features according to three strategies. The first strategy selects for each
	drug separately, the 50 features that are not in the pathway of the drug with
	the largest variance and the 50 features that are in the pathway of the drug
	with the largest variance. If less than 50 features that are in the pathway 
	are available, we select the maximum number available. This strategy ensures
	that the features are somewhat balanced with respect to the external feature
	covariate, i.e., pathway membership. The second strategy selects the 
	100 features with the largest variance, irrespective of pathway membership,
	while the last stategy selects 1000 features with the largest variance,
	again irrespective of pathway membership. These three strategies lead to
	three different analysis outcomes.

	To compute performance measures, we randomly split the data into training and
	test sets, containing 
	241 cell lines each, estimate the models on the training data and compute 
	performance measures on the test data. We repeated this procedure 50 times
	and average performance measures over the splits.
	The estimated 
	models are: four models like in (\ref{eq:prior}) but with different external 
	variables, named NIG1, NIG2, NIG3, and NIG4 in the following. NIG1 estimates
	only one overall $\phi$ and one overall $\chi$, NIG2 uses the external 
	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
	fixed. In addition, we estimate regular lasso and ridge models.

@@

res1 <- 
	\section{Discussion}\label{sec:discussion}
	The preceding presents a novel model for drug sensitivity
	prediction from a set of high dimensional molecular features. The model allows
	for the inclusion of discrete and continuous external covariates on both the 
	drugs and features. Inclusion of the external information is through 
	data-driven and adaptive empirical Bayes estimation of the hyperparameters in 
	the normal inverse Gaussian prior model (\ref{eq:prior}). Estimation is 
	efficient and scales well with the number of features and samples.

>>>>

	We estimate four models: (i) an NIG model that does include external 
	covariates in the estimation called NIG1 in the following, (ii) the NIG model
	estimated as in Section \ref{sec:estimation} called NIG2, and regular (iii)
	lasso and (iv) ridge models.
	Exclusion of the external covariates as in the NIG1 model
	amounts to direct estimation of one common expected 
	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
	$\phi_{jd}$ and/or $\chi_d$ as in the NIG2 model. All 
	models are fit on 241 randomly selected cell lines, while performance measures 
	are estimated from the remaining 241 cell lines. We repeat the whole
	simulation 50 times.

@@

	Table \ref{tab:simulations_gdsc_tab1} displays several median performance
	measures for simulation version 1. The table includes as a reference 
	the null model, which
	is an empty model that only fits the means of the $\y_{d}$.
	The median evidence lower bounds (ELBO) approximates the marginal
	likelihoods of models NIG1 and NIG2 on both training and test data, while 
	estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
	performance. Focussing on MSE, we see that the NIG2 model, that includes
	the external covariates, performs best, both in terms of estimation and 
	prediction. Model fit is best for the NIG1 model that does not include the
	external covariates, according to the ELBO, calculated on the test data.
	The lower MSEs confirm that the model that
	includes the external data (NIG2) learns the underlying structure 
	in the data better than model NIG1 that does not include the external 
	data.
<<simulations_gdsc_tab1>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 1, with the best performing model in bold.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

@@

	We expect that drugs that have been clinically approved are easiest to 
	predict and hence yield the largest prior $\beta_{jd}$ variances, followed by
	the drugs in clinical development, and the experimental drugs. Likewise we 
	expect the targeted drugs to yield the largest prior $\beta_{jd}$ variances, 
	followed by the cytotoxic drugs, and the unkown target drugs. For the
	feature covariate, we expect that genes that are in the pathway of the drug
	are more predictive than genes that are not, i.e., 
	they have larger prior $\beta_{jd}$ variances than drugs that are not in 
	the pathway. Note that large $\beta_{jd}$ variances translate to large
	prior $\gamma_{jd}^2$ and $\tau_d^2$ means.

	\subsection{Results}
	As a preliminary step, we select
	features according to three strategies. The first strategy selects for each
	drug separately 50 features that are not in the pathway of the drug with
	maximum variance and 50 features that are in the pathway of the drug
	with maximum variance. If less than 50 features in the pathway are available,
	we select all available features. This strategy ensures
	that the features are somewhat balanced with respect to the external feature
	covariate, i.e., pathway membership. The second strategy selects the 
	100 features with the largest variance, irrespective of pathway membership,
	while the last stategy selects 1000 features with the largest variance,
	again irrespective of pathway membership. These three strategies lead to
	three different analyses described below.

	To compute performance measures, we randomly split the data into training and
	test sets, containing 
	241 cell lines each, estimate the models on the training data and compute 
	performance measures on the test data. We repeated this procedure 50 times
	and average performance measures over the splits.
	The estimated 
	models are: four models like in (\ref{eq:prior}) but with different external 
	variables, named NIG1, NIG2, NIG3, and NIG4 in the following. NIG1 does 
	not use the external covariates and estimates
	one overall $\phi$ and one overall $\chi$, NIG2 uses the external 
	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
	fixed. In addition, we estimate regular lasso and ridge models. The null model
	that fits the means of the $\y_d$ is included as a reference.

@@

<<analysis_gdsc_fit3>>=
fit <- read.table("results/analysis_gdsc_fit3.txt", row.names=NULL)
temp <- fit[, 1]
fit <- as.matrix(fit[, -1])
rownames(fit) <- temp

res <- read.table("results/analysis_gdsc_res3.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

<<analysis_gdsc_est3_feat>>=
options(knitr.kable.NA="", digits=3)
kable(est1,
      col.names=colnames(est1), align="r",
      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data for feature selection strategy 3.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<analysis_gdsc_est3_drug>>=
kable(est2,
      col.names=colnames(est2), align="r",
      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data for feature selection strategy 3.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<analysis_gdsc_tab3>>=
kable(tab,
      col.names=colnames(tab), align="r",
      caption="Performance measures from GDSC data for feature selection strategy 3.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

	\section{Discussion}\label{sec:discussion}
	The preceding presents a novel model for drug sensitivity
	prediction from a set of high dimensional molecular features. The model allows
	for the inclusion of discrete and continuous external covariates on both the 
	drugs and features. Inclusion of the external information is through 
	data-driven and adaptive empirical Bayes estimation of the hyperparameters in 
	the normal inverse Gaussian prior model (\ref{eq:prior}). Estimation is 
	efficient and scales well with the number of features and samples.
